---
title: "Report"
format: pdf
editor: visual
---

```{r}
#| include: false
source(here::here("scripts", "data_analysis.R"))
```

## Introduction

In this paper I argue that Investor-State Dispute Settlements are an institution which further deepens the asymmetric power dynamics inherent in the international economic order. These ISDS allow investors (mostly private companies) to sue states for alleged violations of investment agreements. On the one hand this is clearly a threat to local autonomy and democratic self-determination, because it gives companies the power to go against any policies that would potentially decrease their profit. There have been various cases where companies sued for their lost profit because these states implemented "green policies" demanded by its citizens. While this in itself is deeply problematic, the focus of this paper is on how ISDS could potentially hinder the development of developing countries by reinforcing existing global economic hierarchies.

This paper builds upon dependency theory as its theoretical framework. The essence of the dependency theory is that the foreign capital that comes from developed capitalist centers, mainly western states, creates an imbalance between them and their periphery (developing states) as well as between the peripheral societies. The profits that are produced in the periphery are appropriated by the capitalist center. This renders the creation of an independent economy in the developing states a hard task, while the developed regions can further consolidate their economic dominance (Frank 1967; Gosh 2019; Tausch 2010). In his study, Tausch (2010) finds that in the long term, higher multinational capital penetration leads to greater inequality and lower performance on standard indicators of development as well as lower predicted economic growth. These findings provide some empirical evidence for the dependency theory.

If dependency theory holds, we would expect ISDS to be used disproportionately by investors from high-income states against developing country respondents. This paper tests this hypothesis empirically using data from the International Center for Settlement of Investment Disputes (ICSID), the world's leading institution for investor-state arbitration.

## Method

### Data Collection

In order to collect the different ISDS cases, I combined a API request with a web scraping approach.

The first step in the data collection process was to collect all case ID's, since they would be later needed to loop through all the case links. Fortunately it was farly easy to do so through a simple undocumented (internal) API request, because although the ICSID does not have a documented API, there webpage is built by calling a internal API. To collect the data, I simply created a GET request to the same API, which returned structured JSON data on all registered cases.

``` r
# make a GET request to the API endpoint
  http_response <- request("https://icsid.worldbank.org/api/all/cases") |>
    req_perform()
```

I made the request and unpacked the response using the `httr2` package.

Since not all the information needed for this analysis were contained in this initial dataset, the second step was to create a web-scraper that extracted the information from every single case internet page. As mentioned above, the web links to these sites consisted of the base-url and the case ID so looping through them was fairly easy. I first created a vector with all the case IDs and then a function that put together the weblink and then went to the website and scraped the required information.

``` r
# defining the base url
base_url <- "https://icsid.worldbank.org/cases/case-database/case-detail?CaseNo="
  

# defining a function to scrape case details
  scrape_detail <- function(case_no) {
    Sys.sleep(0.2)  # wait 0.2 seconds after every request to avoid overwhelming the server
    
    url  <- paste0(base_url,case_no) # construct the url for the case details page using the case number
    
    # tryCatch catches the error before it crashes the loop
    page <- tryCatch(
      read_html(url),
      error = function(e) {
        message("Failed: ", case_no, " — ", e$message)
        return(NULL)
      }
```

The first couple of times I ran the function, it failed and then stopped so I created a second part (not displayed above) which would catch errors before the function would crash and just fill everything in NA.

Within this function, there is a second function (`get_field)` which parses the html code for the required information and builds a single row out of this information. When running the function for all the case IDs

``` r
case_details <- map_dfr(case_nos, scrape_detail, .progress = TRUE)
```

the `map_dfr` directly puts the row from each case ID directly into a data frame, so that I don't end up with hundreds of single tibbles that each only contains one row. With that the data collection process is already completed.

### Data Cleaning

The first step of the data cleaning was to make one dataset out of these two raw data collections. To do so, I basically had to extract the information from the JSON file and combine it with the data frame from the web scraping. I did so by creating a function:

``` r
cases_df <- map_df(cases_list, function(case) { 
    
  # extract the first element of the caseproceedings list, which contains the date_registered information
    proc <- case$caseproceedings[[1]]
    
    tibble(
      case_no         = case$caseno,
      claimant        = case$claimant,
      respondent      = case$respondent,
      date_registered = as.Date(mdy_hms(proc$dateregistered)),
      status          = case$status
    )
  })
```

The function is similar to the one I used for the web scraping. I first looked where the important information were stored and then created a tibble for each case that contained all these extracted information. It was then again mapped to a data frame.

In a second step I had to do some data cleaning and because for my analysis I needed the countries to be stored in a standardized format. For the respondent countries this was a straight-forward task, since almost all of the respondents were states and were stored as such. For the few exceptions where the respondents were not a state, a replaced the value with NA because these cases are not of interest. To convert the states to iso3c standardized format I used the r-package `countrycode`, which takes different types of input (in this case `state.name`) and gives back a defined output (in this case `iso3c`). There were only very little exceptions for which I had to match the state to the iso3 format manually.

For the claimants this task was substantially harder, because only the nationality of the sueing company was written down in brackets. One one hand this made the `countrycode` packages useless and it furthermore required the extraction of the nationality from the entire variable field.\
Since the nationality was in brackets, I could just use a regular expression to extract it. The solution for the second problem required the use of another database, which I found on Github. This table stored each state and how the nationality was called and even the iso3c notation. So downloaded that table and wrote a function that matched the nationality from my database with the nationality of the downloaded github table and extracted the corresponding iso3c notation. However since not all the nationalities in my dataset were stored as nationalities but some as countries, I created a fallback, that if in case the nationality did not match, it should check the country name.

``` r
  # Try to match claimant company to a country
    icsid_df <- icsid_df |>
      mutate(
        iso3_claimant = coalesce(
          # Step 1: match against nationality adjectives (e.g. "Dutch", "French")
          nationality_lookup$iso3[match(claimant_nationality, nationality_lookup$Nationality)],
          # Step 2: fall back to country name match (e.g. "Luxembourg", "Seychelles")
          nationality_lookup$iso3[match(claimant_nationality, nationality_lookup$Name)]
        )
      )
```

Unfortunately, this method was not as reliable as the one that worked with the respondents, so I needed to match various cases manually. A second problem that occurred, was that some cases had more than one claimant. While this is probably not the statistically best solution, I ended up creating multiple rows, for each claimant one, for the cases with two or more claimants.

Furthermore, since I was not sure if I wanted to do some grouping by region later on in the analysis, I decided to already integrate for the World Bank region and UN subregion for both the claimant and respondent country into the dataset during the data cleaning process. This was a easy process, since the `countrycode` package contains these regions.

Last but not least, I needed to classify countries according to their economic status. Several approaches exist for this purpose, such as GDP per capita, the Human Development Index, or simple continent-based groupings. I decided to use the World Bank income classification because it is a widely recognized and standardized measure used extensively in international economics and development research, making it particularly well suited for testing hypotheses derived from dependency theory. It classifies countries into four categories based on Gross National Income (GNI) per capita: Low income, Lower middle income, Upper middle income, and High income. These four categories map naturally onto the core-periphery framework of dependency theory, with High income countries representing the core and Low and Lower middle income countries representing the periphery. GNI is more appropriate than GDP for measuring a country's wealth because it captures income flows across borders, which is particularly relevant in the context of dependency theory where profit extraction from the periphery to the core is a central concept.

The `wbstats` package directly uses the World Banks API, so using this package instead of another API request saved me a lot of work and made the process less complicated.

``` r
    # Load World Bank income category data
    income_df <- wb_countries() |>
      select(iso3c, income_level)
    
    # add the columns to the df
    icsid_df <- icsid_df |>
      left_join(income_df |> rename(income_claimant  = income_level),
                by = c("iso3_claimant"  = "iso3c")) |>
      left_join(income_df |> rename(income_respondent = income_level),
                by = c("iso3_respondent" = "iso3c"))
```

First, a additional data frame, that contained the iso3c and the income classification from the world bank was created, using the `wbstats` package. By matching the `iso3_claimant` and `iso3_respondent` variables in my ICSID dataset against the ISO3 codes in the World Bank income data frame looking up the corresponding income group, the two new variables were created.

## Analysis and Results

After having collected all the necessary data and having compiled them into a clean data frame, I was able to analyze it.

### Claimant vs. Respondent Countries

I started the analysis by plotting the number of how many cases each country was involved in an ICSID case as a claimant and as a respondent. In @fig-map. The North-South divide is clearly visible, with European countries and the US clearly being more involved in this disputes as claimants and developing countries, especially in Latin America displaying a much larger number in the respondent position.

```{r}
#| label: fig-map
#| echo: false
#| fig-cap: "ICSID Cases by Claimant and Respondent Country"
#| fig-pos: "H"


p_cases_map
```

The next step to investigate the hypothesis and to quantify that result was to check wether there is a correlation between the number of cases the different income groups were suing each other. As mentioned above, to do so, I used the income classification for each country from the world bank. Not included in this report, there is a plot, which shows the income classification for each country.

I used a chi-squared test to see if there is any correlation. With a p-value of 2.2e-16 the test shows that there is a significant correlation between the income group of the claimant and the income group of the respondent. This means that there is a pattern and that it is not random whom sues who. In @fig-heatmap-income these patterns are made visible. What can be drawn from this plot is that ICSID cases are predominantly a phenomenon among wealthier countries. However, only based on the heatmap it is not possible to say anything about the statistical significance.

```{r}
#| label: fig-heatmap
#| echo: false
#| fig-cap: "ICSID Cases by Income Group"
#| fig-pos: "H"


p_heatmap
```

When looking at the standardized residuals from the chi-square test we find that disputes among claimants and respondents are over represented (greater than expected). By contrast, high-income claimants suing lower-income respondents is either neutral or negative across all cells, suggesting no disproportionate targeting of poorer states by wealthier investors.\
While these findings suggest that in general, high-income claimants don't disproportionately sue poorer states *overall,* this still might be the case when looking at specific sectors.

In a second step I therefor classify the different sectors into extractive sectors and non-extractive sectors with the first containing sectors such as oil, water, agriculture and forestry because these sectors are characterised by the exploitation of natural resources that are geographically fixed making them particularly prone to disputes between foreign investors and host states. The latter group contains disputes regarding the financial sector, information and communication sector and the energy sector.

@fig-heatmap-sector shows that in extracive sectors, high income claimants disproportionally sue upper middle income countries.

```{r}
#| label: fig-heatmap-sector
#| echo: false
#| fig-cap: "ICSID Cases by Sector and Income"
#| fig-pos: "H"

p_sector_heatmap

```

Again it is necessary to not only look at the distribution but for the statistical findings as well. In the extractive sector we find partial supports the argument that ICSID cases are a tool used by wealthy nations to protect the interests of their investors in resource-rich emerging economies. However, the effect is concentrated against upper-middle-income states not the poorest countries. ave historically pursued nationalisations or regulatory interventions in the extractive sector. This suggests that investment arbitration in the extractive sector functions less as a mechanism of exploitation in the classical sense, and more as a tool for wealthy investors to challenge state sovereignty over natural resources.
